---
title: "Learning Progress"
---

Below is a record of R exercises I've worked through, shown in the order completed. Each entry includes what I was trying to solve and the code I used to solve it.

```{r setup, include=FALSE}
df1 <- data.frame(
  id = c("1", " 2 ", " 3 ", "4","5","6"," 7 ","8","9","10"),
  age = c(25, NA, 31, 28, 40, NA, 22, 35, 29, NA),
  income = c(50000, 60000, NA, 65000, 72000, 68000, NA, 74000, 69000, 71000)
)

survey <- data.frame(
  RespondentID = c("R1","R2","R2","R4","R5","R6","R7","R8","R9", NA,"R10","R11","R12","R13","R14","R15","R16","R17","R18", NA),
  Age_Group = c("18-24","25-34","25-34","35-44","18 - 24","45 - 54","18-24","25-34","35-44","25-34","18-24","25-34","25-34","35-44","18 - 24","45 - 54","18-24","25-34","35-44","25-34"),
  Satisfaction = c("5","4","4","3","5", NA, "2","4","3","4","5","4","4","3","5", NA, "2","4","3","4"),
  Would_Recommend = c("yes","YES"," YES  ","no ","YES ","no  ","YES","no", NA,"yes ","yes","YES"," YES ","NO","No","no","YES","no", NA,"yes "),
  stringsAsFactors = FALSE
)

library(tidyverse)
```

------------------------------------------------------------------------

## Entry 1: Survey Data Cleaning: Duplicates, Case, Spacing, and Type Conversion

**Task:** This survey dataset had several common real-world data issues at once: duplicate respondent rows, inconsistent capitalization and spacing in text responses, a numeric rating stored as character text, and missing respondent IDs. I worked through each issue systematically, checking the data before and after each fix, using both base R and tidyverse approaches.

```{r}
# preview the raw, uncleaned data
survey
```

### Duplicates

Before removing anything, I first confirmed whether repeated IDs were true duplicates or just similar looking rows, including checking whether missing values were adding noise to that comparison.

```{r}
# view all rows sharing the same RespondentID
survey[survey$RespondentID == "R2", ]

# exclude NA rows to see only the real comparison
survey[survey$RespondentID == "R2" & !is.na(survey$RespondentID), ]

# systematically check the whole dataset for duplicate rows
duplicated(survey)

# remove the duplicates
survey <- survey[!duplicated(survey), ]

# verify it worked
nrow(survey)
survey[survey$RespondentID == "R2" & !is.na(survey$RespondentID), ]
```

### Would_Recommend: fixing case and spacing

```{r}
# check inconsistencies (base R)
table(survey$Would_Recommend)

# base R fix
survey$Would_Recommend <- tolower(trimws(survey$Would_Recommend))
table(survey$Would_Recommend)
```

```{r}
# tidyverse fix
survey$Would_Recommend <- str_squish(survey$Would_Recommend)
survey$Would_Recommend <- str_to_lower(survey$Would_Recommend)
table(survey$Would_Recommend)

# tidyverse piped version
survey$Would_Recommend <- survey$Would_Recommend %>%
  str_squish() %>%
  str_to_lower()
table(survey$Would_Recommend)
```

### Age_Group: fixing spacing around hyphens

This one took two passes, my first attempt with `str_squish()` and `str_trim()` alone didn't fully resolve the inconsistent spacing around hyphens, which is a useful reminder to always re-check output rather than assume a fix worked.

```{r}
# check inconsistencies
table(survey$Age_Group)

# base R fix
survey$Age_Group <- gsub(" - ", "-", survey$Age_Group)
table(survey$Age_Group)
```

```{r}
# first stringr attempt, didn't fully fix spacing
survey$Age_Group <- str_squish(survey$Age_Group)
survey$Age_Group <- str_trim(survey$Age_Group)
table(survey$Age_Group)

# targeted fix using str_replace_all()
survey$Age_Group <- str_replace_all(survey$Age_Group, " - ", "-")
table(survey$Age_Group)

# tidyverse piped version
survey$Age_Group <- survey$Age_Group %>%
  str_replace_all(" - ", "-")
table(survey$Age_Group)
```

### Satisfaction: converting text to numeric

```{r}
# check current type and values
str(survey$Satisfaction)
table(survey$Satisfaction)

# base R fix
survey$Satisfaction <- as.numeric(survey$Satisfaction)
str(survey$Satisfaction)
table(survey$Satisfaction)
```

```{r}
# tidyverse fix
survey$Satisfaction <- survey$Satisfaction %>%
  as.numeric()
str(survey$Satisfaction)
```

### RespondentID: handling missing values

```{r}
# check how many are missing, and view those rows
sum(is.na(survey$RespondentID))
survey[is.na(survey$RespondentID), ]

# decision: label unknown IDs rather than drop the rows
survey$RespondentID[is.na(survey$RespondentID)] <- "R_unknown_1"

# verify it worked
sum(is.na(survey$RespondentID))
survey[survey$RespondentID == "R_unknown_1", ]
```

**What this showed me:** Cleaning rarely works on the first attempt, the `Age_Group` column looked fixed after `str_squish()` and `str_trim()`, but checking the output revealed it wasn't. Labeling the missing `RespondentID`s instead of dropping them was a deliberate choice to preserve those rows for analysis, but it's a decision that needs to be documented as a limitation in any later write-up, since those responses aren't truly identifiable.

------------------------------------------------------------------------

## Entry 2: Mean Imputation: Choosing a Method Based on Distribution Shape

```{r}
# a small practice dataset with some missing values in age and income
df1
```

**Task:** Before filling in missing values, I wanted to understand whether mean or median imputation was the right choice, not just apply one automatically. That meant checking the shape of the distribution first, comparing the mean and median to see how close they were, and then actually watching what happened to the distribution after imputing.

```{r}
# check for missing values
sum(is.na(df1$age))
df1[is.na(df1$age), ]

# check the shape of the distribution before imputing anything
hist(df1$age, main = "Distribution of Age", xlab = "Age")

# compare mean and median
mean(df1$age, na.rm = TRUE)
median(df1$age, na.rm = TRUE)

# mean (30) and median (29) were close, with no major outlier or skew,
# so mean imputation was a reasonable, defensible choice here

# apply mean imputation
df1$age[is.na(df1$age)] <- mean(df1$age, na.rm = TRUE)

# verify it worked
sum(is.na(df1$age))

# check the shape again after imputing, to see the effect
hist(df1$age, main = "Distribution of Age After Mean Imputation", xlab = "Age")
```

**What this showed me:** After filling in the missing values with the mean, the 25 to 30 bin gained extra values and became a visible peak, shifting the histogram from a fairly flat shape toward a single-peaked one. With only 10 data points, imputation didn't just fill gaps, it visibly changed the shape of the distribution. That's an important caveat to carry into any writeup that uses imputed data, especially with small samples.

## Same Task, Tidyverse Version

To compare approaches, I repeated the same imputation using tidyverse syntax instead of base R.

```{r}
# reset a fresh copy of the data (since df1 was already modified above)
df1_tidy <- data.frame(
  age = c(25, NA, 31, 28, 40, NA, 22, 35, 29, NA)
)

# apply mean imputation using dplyr's mutate and ifelse
df1_tidy <- df1_tidy %>%
  mutate(age = ifelse(is.na(age), mean(age, na.rm = TRUE), age))

# verify it worked
sum(is.na(df1_tidy$age))
df1_tidy$age
```

Both approaches produce the same result: base R modifies the column directly with `[is.na(...)] <-`, while tidyverse expresses the same logic inside `mutate()` with `ifelse()`, as part of a pipeline. The tidyverse version reads more like a sentence describing the transformation, while base R is more direct indexing.
