---
title: "Learning Progress"
---

Below is a record of R exercises I've worked through, shown in the order completed.I worked through each issue systematically, checking the data before and after each fix, using both base R and tidyverse approaches.

```{r setup, include=FALSE}
df2 <- data.frame(
  student_id = 101:110,
  name = c("Ava", "Noah", NA, "mia", "Leo", "Zoe", NA, "Eli", "chloe", "sam"),
  final_score = c(88, NA, 92, 75, 81, NA, 84, 90, NA, 86),
  passed = c("yes", "no", NA, "yes", "yes", "no", NA, "yes", "no", "yes"),
  stringsAsFactors = FALSE
)

library(tidyverse)
```

------------------------------------------------------------------------

## Entry 1: df2 Data Cleaning: Case, Spacing, and Missing Data

**Task:** This df2 dataset had several common real-world data issues at once:inconsistent capitalization, and spacing at data entry and, missing respondent data.

```{r}
# preview the raw, uncleaned data
df2
```

### Cleaning: "Name" Column for Case, Spacing, and, Missing Data

Before cleaning anything, I first inspected the "name" column for case and spacing inconsistencies,then removed any occurrences. similarly, I used the same approach after having observed the missing values.those occurrences were excluded from the data.

```{r}
#checked for case and space inconsistencies
unique(trimws(tolower(df2$name)))   

#converted all names to title case
df2$name <- str_to_title(df2$name) 

#removed white spacing
df2$name <- trimws(df2$name) 

#Verified that it worked
unique(df2$name)

#removed missing values
df2_cleanone <- df2[!is.na(df2$name), ]

#verified that it worked
df2_cleanone
```

### Cleaning: "final_score" Column for Distinct Values, and Missing data

When cleaning the "final_score" column I did not check for case or space inconsistencies as I used natural visual inspection and determined to only view the uniqueness of the data and to remove the missing values.

```{r}
#viewed the distinct values
unique(df2$final_score) 

#removed missing values
df2_cleanthree <- df2[!is.na(df2$final_score), ]

#verified that it worked
df2_cleanthree
```

### Cleaning: "passed" Column for Case, Spacing, and, Missing Data

Before cleaning anything, I first inspected the "passed" column for case and spacing inconsistencies,then removed any occurrences. similarly, I used the same approach after having observed the missing values.those occurrences were excluded from the data.

```{r}
#checked for case and spacing inconsistencies
unique(trimws(tolower(df2$passed)))   

#converted all names to title case
df2$passed <- str_to_title(df2$passed) 

# removed white spacing
df2$passed <- trimws(df2$passed) 

#removed missing values
df2_clpass <- df2[!is.na(df2$passed), ] 

#Verified that it worked
df2_clpass
```

## **What this showed me:**

Data cleaning is a crucial part of working with any dataset, and it should be done thoughtfully to support stronger analysis later. Even though this was only a practice dataset focused on systematically cleaning each column, it also made me think more carefully about how I would analyze the data afterward and what story it might tell.

## **Same Task:** Tidyverse Version

To compare approaches, I repeated the same exercise using tidyverse syntax instead of base R.

------------------------------------------------------------------------

## Entry 2: df2 Data Cleaning: Case, Spacing, and Missing Data

**Task:** This df2 dataset had several common real-world data issues at once:inconsistent capitalization, and spacing at data entry and, missing respondent data.

### Cleaning: "Name" Column for Case, Spacing, and, Missing Data

Before cleaning anything, I first inspected the "name" column for case and spacing inconsistencies,then removed any occurrences. similarly, I used the same approach after having observed the missing values.those occurrences were excluded from the data.

```{r}
#checked for case and space inconsistencies
unique(trimws(tolower(df2$name)))  

#corrected case and spacing
df2$name <- df2$name %>% 
  trimws() %>%
  str_to_title()

#verify that it worked
unique(df2$name)

#removed missing data
df2_cleantwo <-df2 %>%
  filter(!is.na(name))

#verified that it worked
df2_cleantwo
```

### Cleaning: "final_score" Column for Distinct Values, and Missing data

When cleaning the "final_score" column I did not check for case or space inconsistencies as I used natural visual inspection and determined to only view remove the missing values.

```{r}
#removed missing values
df2_cleanthree <- df2 %>%
  filter(!is.na(final_score))

#verify that it worked
df2_cleanthree
```

### Cleaning: "passed" Column for Case, Spacing, and, Missing Data

Before cleaning anything, I first inspected the "passed" column for case and spacing inconsistencies,then removed any occurrences. similarly, I used the same approach after having observed the missing values.those occurrences were also excluded from the data.

```{r}
# checked for case and space inconsistencies
unique(trimws(tolower(df2$passed)))    

#removed missing values
df2_clpass <- df2 %>%
  mutate(passed = trimws(passed) %>% str_to_title()) %>%
  filter(!is.na(passed))
```
